---
title: "Machine Learning Project"
date: "January 28, 2016"
output: html_document
---
#### Human Activity Recognition (http://groupware.les.inf.puc-rio.br/har)

In this exercise I am going to predict the manner in which the people we have data for 
did the exercise they ended up doing.  The variable that indicates this exercise is 
called "classe" in the training set and that is what I will be predicting.
The data in this analysis all came from source: http://groupware.les.inf.puc-rio.br/har.
who have been very generous in allowing their data to be used for this kind of exercise.
I will be using use the other variables (other than "classe") to predict with. 
This is my report describing how I built my model, and how Iused cross validation, 
what I think the expected out of sample error is, and why I made the choices I did. 
I will also use my prediction model to predict 20 different test cases.

Reproducibility
Due to security concerns with the exchange of R code, my code will not be run during the 
evaluation by you, my classmates. You will be be able to view the 
compiled HTML version of my analysis which I will publish under Rpubs and provide a link.

required packages: lattice and ggplot2

### Load my libraries
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(randomForest)
```

### Load the Data
Setting a seed number for Reproducibility
Load the Data for analysis

```{r}
set.seed(54321)

# 
# Pull the Training data into the trainData variable in memory
#
trainData <- read.table("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                        header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""), dec=".", strip.white=TRUE)

dim(trainData)

#
# Pull the Test data into the testData variable in memory
#
testData <- read.table("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                       header=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""), dec=".", strip.white=TRUE)
dim(testData)

```

### Cleaning the Data
Now I want to cleanup the near zero variance
Get rid of the near zero variances as they will cause poor results in my final
prediction model.

```{r}
trainingNZV <- nearZeroVar(trainData, saveMetrics=TRUE)
trainingCleanup <- trainData[!trainingNZV$nzv]
dim(trainingCleanup)
```

Get rid of the first column.  X is a sequential number for each row so not really going to help 
with predicting anything.

```{r}
trainingCleanup <- trainingCleanup[c(-1)]
```

Now, i want to get rid of variables with too many Null values. 
For Variables that have more than a 30% threshold of Null values; I’m going to leave them out:

```{r}
trainingFinal <- trainingCleanup
for(a in 1:length(trainingCleanup)) {
  if(sum(is.na(trainingCleanup[,a]))/nrow(trainingCleanup) >= .3 ) { 
    for(b in 1:length(trainingFinal)) {
      if(length(grep(names(trainingCleanup[a]), names(trainingFinal)[b]))==1) { 
        trainingFinal <- trainingFinal[,-b] 
      }   
    } 
  }
}

dim(trainingFinal)
```

### Partition the data for model fitting
Now, I want to partition the trainData into a training set and a testing set
for analysis and formula fitting.

```{r}
myTrainingPartition <- createDataPartition(y=trainingFinal$classe, p=0.60, list=FALSE)
myTraining <- trainingFinal[myTrainingPartition, ]
myTesting <- trainingFinal[-myTrainingPartition, ]

dim(myTraining) 
dim(myTesting)
```

I will now take the columns from myTraining that I am going to use
in the model and only keep those columns from the testData that match.

```{r}
clean <- colnames(myTraining[, -58]) 
testData <- testData[clean]
```

Making sure to sync with the same class between myTraining and the testData

```{r}
testData <- rbind(myTraining[2, -58] , testData) 
testData <- testData[-1,]
```

### Finding the best fit prediction model
Trying the ML algorithms for prediction: Decision Tree

```{r}
fit.dt <- rpart(classe ~ ., data=myTraining, method="class")
```

Predicting in-sample error

```{r}
predictions.dt <- predict(fit.dt, myTesting, type = "class")
```

Here is the confusionMatrix to see the results of Decision Tree

```{r}
confusionMatrix(predictions.dt, myTesting$classe)
```

Using ML algorithms for prediction: Random Forests

```{r}
fit.rf <- randomForest(classe ~. , data=myTraining)
```

Predicting in-sample error

```{r}
predictions.rf <- predict(fit.rf, myTesting, type = "class")
```

Here is the confusionMatrix to see the results of Random Forest

```{r}
confusionMatrix(predictions.rf, myTesting$classe)
```
 
```{r}
plot(fit.rf)
```

Plot showing the Mean Decrease Gini values indicating those predictors that are stronger 
(with the lower values)

```{r}
varImpPlot(fit.rf)
```

### Choice of best fit is:
Random Forests is a better fit for predicting in this case
Here is the prediction 
Random Forests gave an Accuracy in the myTesting dataset of 99.87% which 
was more accurate that from the Decision Tree. 

The expected out-of-sample error is 100-99.87 = 0.13%.

Since the sample error is low, this suggests to me that my prediction model is accurate
and therefore, a good fit.

### Predictions of test data
```{r}
predictions.rf.test <- predict(fit.rf, testData, type = "class")
predictions.rf.test
```
